---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# grantpubcite

<!-- badges: start -->
<!-- badges: end -->

The 'grantpubcite' package is used to query the [NIH Reporter][]
database for funded grants and, optionally, publications associated
with grants in NIH Reporter. The citation history of publications can
discovered using [iCite][].  The 'grantpubcite' package help pages and
resources make extensive use of 'tidyverse' concepts.

[NIH Reporter]: https://reporter.nih.gov/
[iCite]: https://icite.od.nih.gov/

Install the development version of grantpubcite from [github][].

[github]: https://github.com/mtmorgan/grantpubcite

```{r, eval = FALSE}
if (!nzchar(system.file(package = "remotes")))
    install.package("remotes", repos = "https://cran.r-project.org")
remotes::install_github("mtmorgan/grantpubcite")
```

Load the library and other packages to be used in the vignette.

```{r setup, message = FALSE}
library(grantpubcite)
```

See

- [Get started][] for an introduction to the package.
- [Case study: ITCR][] for an illustration of package use analysing
  grants in the NIH Information Technology in Cancer Research
  portfolio of grants. The ITCR is particularly interesting because of
  it's focus on 'dry-lab' technology / software, multi-year duration,
  and sequence of grants representing a progression from small grants
  emphasizing innovation to large grants to sustain essential
  resources.
- [Case study: IOTN][] looks at the NIH NCI Immuno-Oncology
  Translation Network portfolio of grants; the IOTN servers as an
  interesting 'wet-lab' comparison for ITCR.
- [Case study: specific projects][] provides a quick explorartion of
  publications and citations arising from two related grants.

[Get started]: articles/grantpubcite.html
[Case study: ITCR]: articles/case_study_itcr.html
[Case study: IOTN]: articles/case_study_iotn.html
[Case study: specific projects]: articles/case_study_project_nums_quick.html

Please note that I am not experienced in this sort of analysis, so any
conclusions should be viewed with great skepticism.
