---
title: "Case study: specific projects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: specific projects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Compiled: `r Sys.Date()`

This article illustrates how one might explore specific project
numbers. Start by attaching our package and others we will use.

```{r setup, message = FALSE}
library(grantpubcite)
library(DT)
```

# Projects

Suppose one is interested in specific projects, and the grant project
numbers are known. Define and search for them.

```{r}
project_nums <- c("U01CA190234", "U24CA194354")
```

Query the NIH Reporter for the projects of interest. The use of
`project_nums=` for the argument required some reflection on the API
schema provided at the end of the [API projects endpoint][project
api].

[project api]: https://api.reporter.nih.gov/#!/Search/Projects

```{r project-numbers}
projects <- reporter_projects(project_nums = project_nums)
```

Explore the data, for instance finding the Funding Opportunity
Announcement the grants were reward under, and the fiscal years in
which they were active.

```{r project-fiscal-years}
projects |>
    group_by(core_project_num) |>
    summarize(
        foa = unique(full_foa),
        first_year = min(fiscal_year),
        last_year = max(fiscal_year),
        project_title = unique(project_title)
    ) |>
    gpc_datatable()
```

# Publications

Use the core project numbers to retrieve known publications

```{r}
core_project_nums <-
    pull(projects, "core_project_num") |>
    unique()
publications <- reporter_publications(core_project_nums = core_project_nums)
```

The number of publications associated with each grant are as follows:

```{r}
publications |>
    count(coreproject) |>
    gpc_datatable()
```

Many publications were attributed to both grants

```{r}
publications |>
    count(pmid, name = "cited_by") |>
    count(cited_by) |>
    gpc_datatable()
```


# Citations

Obtain citation information (including information about the orginal
publication) using the `publications` tibble

```{r}
citations <- icite(publications)
```

There are `r nrow(citations)` citation records, one for each
publication. Explore these interactively

```{r}
citations |>
    select(pmid, citation_count, relative_citation_ratio, year, title) |>
    gpc_datatable()
```

Use a 'join' between tables to associate publications and citations

```{r}
left_join(publications, citations) |>
    group_by(coreproject, year) |>
    summarize(
        n_publications = n(),
        total_citations = sum(citation_count)
    ) |>
    gpc_datatable()
```

# Session information

```{r}
sessionInfo()
```
