---
title: "Case study: ITCR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: ITCR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
datatable <-
    function(x, ...)
{
    if (interactive()) {
        x
    } else {
        DT::datatable(x, ..., rownames = FALSE, options = list(pageLength = 5))
    }
}
```

<!-- badges: start -->
<!-- badges: end -->

# Summary

Note that this is way outside my area of expertise, so I have
undoubtedly made terrible blunders.

The 'grantpubcite' package can be used to query the [NIH Reporter][]
database for funded grants and, optionally, publications associated
with grants in NIH Reporter. The citation history of publications can
discovered using [iCite][].

As a case study, suppose we are interested in grants funded under the
Information Technology in Cancer Research ([ITCR][]) program.

[ITCR]: https://itcr.cancer.gov/

The [Project information][] section shows that I found 87 projects (a
useful sanity check?). The first figure shows that projects funded for
more fiscal years received more funding (not too surprising!). At this
stage and subsequently I found it useful, using the datatables on the
web page, to search for ‘my’ project (U24CA180996, Cancer Genomics:
Integrative and Scalable Solutions in R/Bioconductor) and compare how
we were doing relative to other projects. Frankly, I did not know
where we stood; now I do.
 
I then focused on projects receiving funding in 2020.
 
The [Maturing projects?][] subsection uses project title and contact PI
to ask whether projects graduate between funding activities. Three
(led by Griffith, Karchin, and Liu) seem to have, all from
‘Early-stage’ to ‘Advanced’. I was a little disappointed that there
were not more smaller projects maturing.
 
In the [Publications][] section, we see a positive relationship
between funding (amount or duration) and publication. Again this is
not surprising, but speaks to the notion that scientific software
projects are playing by the same rules as ‘wet lab’ projects; it would
be interesting to identify a comparable wet-lab program for
comparison. There are some outliers, including one project with a
surprisingly large number of publications for the period of the grant,
as well as projects with only a handful of publications even after
multiple years of funding. The [ITCR collaboration][] section shows
that 5 projects collaborated closely enough with each other to be
acknowledged in the same publication.
 
The figure in the [Citations][] section shows what I imagine is a
pretty typical pattern, with most publications having moderate (<100)
citations, and a few having many citations. Scanning the datatable
immediately below the figure indicates that the highly cited
publications are biologically driven, where the paper authors have
presumably cited the relevant software; the three most-cited
publications are all from U24CA180922, which funds the Trinity
software. When I look at the five most highly cited of ‘my’ project
publications (U24CA180996), they are more modestly successful (181 to
337 citations); interestingly, 4 of the 5 publications emphasize
software per se, rather than strictly biological insight.
 
The [ITCR collaborations][] subsection takes a second look at
collaboration, in the sense of ITCR projects citing the works of one
another. 37 projects cite work of 22 other projects.

[NIH Reporter]: https://reporter.nih.gov/
[iCite]: https://icite.od.nih.gov/

# Getting started

See the [Introduction to 'grantpubcite'][grantpubcite] article for
installation, basic use, and a brief introduction to 'tidyverse'
operations.

[grantpubcite]: articles/grantpubcite.html

Load the library and other packages to be used in this article.

```{r setup, message = FALSE}
library(grantpubcite)
library(ggplot2)
library(DT)
```

# Project information

The relevant funding opportunity announcements (FOA) are summarized in
a `tibble` (the 'tidy' version of `data.frame`)

```{r}
foas <- tribble(
    ~full_foa,    ~reissue_of,  ~foa_tag,      ~description,
    ## current FOAs
    "PAR-15-331", "PAR-13-294", "Advanced",    "Advanced Development of Informatics Technologies for Cancer Research and Management (U24)",
    "PAR-15-332", "PAR-12-288", "Early-Stage", "Early-Stage Development of Informatics Technologies for Cancer Research and Management (U01)",
    "PAR-15-333", NA,           "Sustained",   "Sustained Support for Informatics Resources for Cancer Research and Management (U24)",
    "PAR-15-334", NA,           "Innovative",  "Development of Innovative Informatics Methods and Algorithms for Cancer Research and Management (R21)",
    ## earlier FOAs
    "PAR-13-294", "PAR-12-287", "Advanced",    "Advanced Development of Informatics Technology (U24)",
    "PAR-12-287", NA,           "Advanced",    "Advanced Development of Informatics Technology (U24)",
    "PAR-12-288", NA,           "Early-Stage", "Early-Stage Development of Informatics Technology (U01)"
    ## FIXME: add 'Revsion of...' FOAs
)
foas <- select(foas, -reissue_of) # drop 'reissue_of' for clarity
```

The NIH Reporter contains quite a bit of information about each
grant. Following the [Get started][] article, we query the
`reporter_projects()` endpoint, restricting fields included in the
return value to a few of particular interest.

[Get started]: articles/grantpubcite.html

```{r}
include_fields <- c(
    "full_foa",
    "core_project_num",
    "fiscal_year",
    "award_amount",
    "contact_pi_name",
    "project_title"
)
```

Retrieve these fields for all projects in NIH reporter associated with
the funding announcements of interest. Perform some minor data
cleaning by removing leading and trailing whitespace from contact PI
names

```{r}
projects <-
    reporter_projects(
        foa = foas$full_foa,
        include_fields = include_fields
    ) |>
    mutate(contact_pi_name = trimws(contact_pi_name))
```

The result is a tidyverse `tibble`, and we use 'tidy' semantics to
explore the data. For instance, although there are `r nrow(projects)`
returned by NIH Reporter, each project is associated with funding over
multiple fiscal years and perhaps mechanisms (e.g., through
administrative supplements).

```{r}
projects |>
    distinct(full_foa, core_project_num) |>
    left_join(foas, by = "full_foa") |>
    datatable()
```

The number of projects funded by each FOA is

```{r}
projects |>
    left_join(foas) |>
    distinct(foa_tag, core_project_num) |>
    count(foa_tag)
```

The number of projects funded per fiscal year is

```{r}
projects |>
    distinct(full_foa, core_project_num, fiscal_year) |>
    count(fiscal_year)
```

A few projects changed names within the same award; for convenience we
create a table of most-recent project titles

```{r}
project_titles_h <-
    projects |>
    select(fiscal_year, core_project_num, project_title) |>
    arrange(desc(fiscal_year)) |>
    filter(!duplicated(core_project_num)) |>
    select(core_project_num, project_title)
```

The `r nrow(project_titles_h)` projects funded by ITCR are

```{r}
datatable(project_titles_h)
```

We update `projects` to use harmonized project titles.

```{r}
projects <-
    projects |>
    select(-project_title) |>
    left_join(project_titles_h, by = "core_project_num")
```

Three project numbers have been funded under more than one FOA.

```{r}
projects |>
    distinct(full_foa, core_project_num) |>
    count(core_project_num, sort =  TRUE) |>
    filter(n > 1L) |>
    left_join(project_titles_h)
```

The following summarizes project funding across years; there are likely many
caveats to this.

```{r}
project_summaries <-
    projects |>
    group_by(core_project_num) |>
    summarize(
        n_years = length(unique(fiscal_year)),
        amount = sum(award_amount)
    ) |>
    left_join(project_titles_h, by = "core_project_num") |>
    arrange(desc(n_years))

project_summaries |>
    ggplot(aes(n_years, amount)) +
    geom_point() + geom_smooth(method = "lm") +
    scale_x_continuous(breaks = seq(1, 10), minor_breaks = NULL) +
    labs(x = "Fiscal years of award", y = "Total funding")
datatable(project_summaries)
```

## Maturing projects?

A unique aspect of the ITCR FOA structure is that it envisions
projects starting at one FOA and 'maturing' to the next FOA, e.g., an
`Early-Stage` U01 matures to an `Advanced` U24. It is not possible to
assess this by tracking a project number through full FOA, because the
project number changes with FOA. Are there projects with the same
contact PI that have received awards from different FOA tags?

```{r}
tagged_pis <-
    projects |>
    distinct(full_foa, core_project_num, contact_pi_name) |>
    left_join(foas, by = "full_foa") |>
    select(foa_tag, contact_pi_name) |>
    distinct()
```

```{r}
maturing_pis <-
    tagged_pis |>
    count(contact_pi_name, sort = TRUE) |>
    filter(n > 1L) |>
    select(-n) |>
    left_join(
        projects |>
        distinct(full_foa, core_project_num, contact_pi_name, project_title),
        multiple = "all"
    ) |>
    left_join(foas, by = "full_foa") |>
    select(contact_pi_name, project_title, foa_tag) |>
    arrange(contact_pi_name, project_title, foa_tag)
DT::datatable(maturing_pis)
```

Direct inspection suggests that three contact PIs (Giffith, Karchin,
Liu) have each seen one project mature. All are from 'Early-stage' to
'Advanced'.

# Publications

Grantees report publications associated with their grants, and this
information can be retrieved from NIH reporter. Queries are formulated
in a way similar to projects, as described on the NIH Reporter
[publication search API][], a rich set of query criteria can be used,
but fields included in the return are strictly limited.

[publication search API]: https://api.reporter.nih.gov/#!/Search/Publications

Projects funded in 2020 are

```{r}
projects_2020 <-
    projects |>
    filter(fiscal_year == "2020") |>
    left_join(project_summaries) |>
    distinct(
        full_foa, core_project_num,
        n_years, amount, project_title
    )
```

There are `r nrow(projects_2020)` projects receiving funds in fiscal
year 2020. Publications associated with these projects are obtained with

```{r}
publications <- reporter_publications(
    core_project_nums = pull(projects_2020, "core_project_num")
) |>
    select(-applid) # we won't use 'applid'
```

There are `r nrow(publications |> distinct(pmid))` distinct
publications (not all published in 2020).

## Project publication and funding

The most prolific projects are

```{r}
publications_by_project <-
    publications |>
    count(coreproject, sort = TRUE) |>
    left_join(
        projects_2020,
        by = c(coreproject = "core_project_num"),
        multiple = "last"
    )
datatable(publications_by_project)
```

An exceptional number of publications are reported by project number
'U01CA239055'. The relationship between publication and funding period
and amount are visualized as follows.

```{r, fig.show="hold", out.width="45%"}
publications_by_project |>
    ggplot(aes(n_years, n)) +
    geom_point() + geom_smooth(method = "lm") +
    scale_x_continuous(breaks = seq(1, 10), minor_breaks = NULL) +
    labs(x = "Fiscal years of award", y = "Number of publications")

publications_by_project |>
    ggplot(aes(amount, n)) +
    geom_point() + geom_smooth(method = "lm") +
    labs(x = "Total funding", y = "Number of publications")
```

## ITCR collaboration

ITCR emphasizes collaboration between funded projects. Are there
examples of collaboration at the level of publication, i.e., `pmid`
associated with more than one project number?

```{r}
collaborative_pmid <-
    publications |>
    count(pmid, sort = TRUE, name = "n_collab") |>
    filter(n_collab > 1)
collaborative_pmid
```

Which projects are collaborating through shared publication?

```{r}
collaborative_pmid |>
    left_join(publications, multiple = "all") |>
    count(coreproject, sort = TRUE, name = "n_collab") |>
    left_join(project_summaries, by = c(coreproject = "core_project_num")) |>
    select(core_project_num = coreproject, n_collab, project_title) |>
    distinct() |>
    filter(n_collab > 1) |>
    datatable()
```

# Citations

Query [iCite][] for all citations to the publications associated with
ITCR projects funded in 2020, restricting output to include fields of
interest.

```{r}
include_fields <- c(
    "pmid", "year", "citation_count", "relative_citation_ratio",
    "year", "title", "authors", "journal", "doi"
)
citations <- icite(publications, include_fields)
citations <- left_join(publications, citations)
```

The `r nrow(citations)` publications have `r pull(citations,
citation_count) |> sum() |> as.integer()` total citations;
`r filter(citations, citation_count == 0) |> nrow()` publications
have not been cited; not surprisingly uncited publications are recent.

```{r}
citations |>
    group_by(year) |>
    summarize(
        n_projects = length(unique(coreproject)),
        n_pub = n(),
        n_uncite = sum(citation_count == 0L),
        citn_count = sum(citation_count)
    ) |>
    DT::datatable()
```

Citations follow a very familiar pattern, with a few
publications cited frequently.

```{r}
citations |>
    filter(citation_count > 0) |>
    ggplot(aes(x = citation_count)) +
    scale_x_log10() +
    geom_density() +
    xlab("Number of citations") + ylab("Number of publications")
```

Highly cited publications are in high-impact journals, and emphasize
science-related results rather than a software tool *per se* -- the
tool has been used in an important study, and the authors of the study
have acknowledged the tool.

```{r}
citations |>
    arrange(desc(citation_count)) |>
    select(coreproject, pmid, year, citation_count, journal, title) |>
    datatable()
```

Citations per project are summarized below.

```{r}
citations |>
    group_by(coreproject) |>
    summarize(
        n_pub = n(),
        n_uncite = sum(citation_count == 0L),
        citn_count = sum(citation_count)
    ) |>
    arrange(desc(citn_count)) |>
    left_join(project_summaries, by = c(coreproject = "core_project_num")) |>
    datatable()
```

## ITCR collaborations

An opportunity for (indirect) collaboration occurs when one project
cites the work of another project. Thus we query [iCite][] for the
publications that cited ITCR publications, and exclude publications
that are not themselves ITCR publications.

```{r}
include_fields <- c("pmid", "cited_by")
cited_by <- icite(publications, include_fields)

## 'cited_by' is a space-delimited character vector of pmid ids; expand these...
cited_by_pmids <-
    pull(cited_by, "cited_by") |>
    lapply(strsplit, "[[:blank:]]") |>
    unlist(recursive = FALSE)

## ...construct a tibble of pmid / cited_by pmid
pmid_cited_by_itcr <-
    tibble(
        pmid = rep(pull(cited_by, "pmid"), lengths(cited_by_pmids)),
        cited_by = unlist(cited_by_pmids) |> as.integer()
    ) |>
    ## and exclude citations by non-ITCR publications
    filter(cited_by %in% pmid)
```

There are `r nrow(pmid_cited_by_itcr)` citations of ITCR publications
by ITCR projects.

```{r}
core_project_citations <-
    pmid_cited_by_itcr |>
    ## map pmid to coreproject
    left_join(publications, multiple = "all") |>
    rename(core_project_num = coreproject) |>
    ## map cited_by to coreproject
    left_join(
        publications,
        by = c(cited_by = "pmid"),
        multiple = "all"
    ) |>
    rename(cited_by_core_project = coreproject)
```

The following summarizes citations at the project level

```{r}
project_project_citations <-
    core_project_citations |>
    count(core_project_num, cited_by = cited_by_core_project, name = "n_citns")
```

Self-citations are

```{r}
project_project_citations |>
    filter(core_project_num == cited_by) |>
    arrange(desc(n_citns)) |>
    left_join(project_summaries) |>
    select(-cited_by) |>
    rename(self_citn = "n_citns") |>
    datatable()
```

One can view collaboration as the number of projects that cite a
particular project...

```{r}
project_project_citations |>
    filter(core_project_num != cited_by) |>
    group_by(core_project_num) |>
    summarize(n_collaborators = length(unique(cited_by))) |>
    arrange(desc(n_collaborators)) |>
    left_join(project_titles_h) |>
    datatable()
```

...or the number of projects a project cites

```{r}
project_project_citations |>
    filter(core_project_num != cited_by) |>
    group_by(cited_by) |>
    summarize(n_collaborators = length(unique(core_project_num))) |>
    arrange(desc(n_collaborators)) |>
    left_join(project_titles_h, by = c(cited_by = "core_project_num")) |>
    datatable()
```

# Session information

```{r}
sessionInfo()
```
