---
title: "Case study: IOTN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: IOTN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Compiled: `r Sys.Date()`

Note that this is way outside my area of expertise, so I have
undoubtedly made terrible blunders.

# Introduction

The 'grantpubcite' package can be used to query the [NIH Reporter][]
database for funded grants, and the publications associated with those
grants. The citation history of publications can be discovered using
[iCite][].

[NIH Reporter]: https://reporter.nih.gov/
[iCite]: https://icite.od.nih.gov/

In this case study we look at the Immuno-Oncology Translational
Network ([IOTN][]), a consortium supported by the [NIH Cancer Moonshot][]
to accelerate translation of basic discoveries to clinical
applications to improve immunotherapy outcomes.

[IOTN]: https://www.iotnmoonshot.org/en/
[NIH Cancer Moonshot]: https://www.cancer.gov/research/key-initiatives/moonshot-cancer-initiative

# Getting started

See the [Introduction to 'grantpubcite'][grantpubcite] article for
installation, basic use, and a brief introduction to 'tidyverse'
operations.

[grantpubcite]: ./grantpubcite.html

Load the library and other packages to be used in this case study.

```{r setup, message = FALSE}
library(grantpubcite)
library(ggplot2)
library(DT)
library(visNetwork)
```

This document is written in Rmarkdown; code chunks used to generate
each table or figure can be shown by toggling the 'Details' widget.

# Projects

Funding Opportunity Announcements for IOTN are as follows:

<details>
```{r}
foas <- case_study_foa_iotn()
```
</details>

```{r, echo = FALSE}
gpc_datatable(foas, truncate_columns = FALSE)
```

Discover projects funded through these FOAs by querying NIH Reporter,
and use the information returned by a the query to summarize the
projects.

<details>
```{r}
project_summary <- program_projects(foas)
```
</details>

```{r, echo = FALSE}
project_summary |>
    gpc_datatable()
```

There are `r nrow(project_summary)` projects. The projects include
large U54, resource sharing and management centers, as well as U01 and
UG3 awards.

<details>
```{r}
plot <-
    project_summary |>
    left_join(foas, by = "full_foa") |>
    ggplot(aes(tag, award_amount)) +
    scale_y_continuous(labels = scales::comma) +
    geom_boxplot(outlier.color = NA) +
    geom_jitter() +
    xlab("Funding type") + ylab("Amount ($)")
```
</details>

```{r echo = FALSE}
plot
```

# Publications and citations

Publications associated with projects are discovered using [NIH
Reporter][]. Citations are from [iCite][].

<details>
```{r}
program_publications <- program_publications(foas)

publications <-
    program_publications |>
    select(-c("full_foa", "core_project_num")) |>
    distinct() |>
    arrange(desc(citation_count))

publications_per_project <-
    program_publications |>
    count(core_project_num, name = "n_pub") |>
    left_join(project_summary, by = "core_project_num") |>
    select(-c("full_foa", "project_title", "fiscal_year")) |>
    arrange(desc(n_pub))

collaboration_summary <-
    program_publications |>
    count(pmid, name = "n_collaborators") |>
    count(n_collaborators, name = "n_publications")
```
</details>

There are `r publications |> NROW()` publications. Project U01DK124165
had a very important publication with PMID 32839624.

```{r, echo = FALSE}
publications |>
    gpc_datatable()
```

About 13% of publications involved collaboration between projects; one
publication represented an extensive collaboration.

```{r, echo = FALSE}
collaboration_summary |>
    gpc_datatable()
```

The network below show collaborations between projects, excluding the
extensive colloration. Hover over nodes to see project number and
title. The width of edges is proportional to the square root of the
number of co-publications; dashed lines indicate a single
copublication.

<details>
```{r}
## pubs <- program_publications(foas)
## pubs |> count(pmid) # pmid 32554617 is highly collaborative
copub_data <- copublication_data(foas, exclude = "32554617")

nodes <-
    copub_data |>
    ## exclude
    tidyr::pivot_longer(dplyr::starts_with("core_project_num")) |>
    distinct(id = value) |>
    left_join(
        project_summary |>
        select(id = "core_project_num", project_title) |>
        distinct(),
        by = "id"
    ) |>
    mutate(
        size = 10,
        title = paste0(id, ": ", .data$project_title)
    ) |>
    arrange(id)

edges <-
    copub_data |>
    mutate(
        from = core_project_num.x,
        to = core_project_num.y,
        width = 3 * sqrt(n),
        smooth = FALSE,
        dashes = n == 1L
    )

network <-
    visNetwork(nodes, edges) |>
    visLayout(randomSeed = 123) |>
    visOptions(highlightNearest = TRUE)
```
</details>

```{r, echo = FALSE}
network
```

The figure below shows publications per year...

<details>
```{r}
plot <-
    publications |>
    ggplot(aes(factor(year))) +
    geom_bar() +
    xlab("Year") + ylab("Number of publications")
```
</details>

```{r, echo = FALSE}
plot
```

...and the distribution of citations per publication

<details>
```{r}
plot <-
    publications |>
    filter(citation_count > 0) |>
    ggplot(aes(citation_count)) +
    scale_x_log10() +
    geom_density() +
    xlab("Number of citations") + ylab("Number of publications")
```
</details>

```{r, echo = FALSE}
plot
```

Project-level citations are summarized by the number of publications,
the total citation count, and the sum of the 'relative citation
index', a measure provided by iCite standardizing the impact of
publications by year and field of study.

<details>
```{r}
citations_by_project <-
    program_publications |>
    group_by(core_project_num) |>
    summarize(
        n_pub = n(),
        n_citn = sum(citation_count),
        total_rcr = sum(relative_citation_ratio, na.rm = TRUE)
    ) |>
    left_join(project_summary, by = "core_project_num") |>
    arrange(desc(total_rcr))
```
</details>

```{r, echo = FALSE}
citations_by_project |>
    gpc_datatable()
```

The relationship between amount of funding and relative citation ratio
is show in the figure below; mousing over points shows the underlying
data and associated core project number. Large 'U54' projects do not
emphasize publication.

<details>
```{r}
plot <-
    citations_by_project |>
    ggplot(
        aes(
            award_amount, total_rcr,
            text = paste0(core_project_num, ": ", gpc_shorten(project_title))
        )
    ) +
    geom_point() +
    scale_x_continuous(labels = scales::comma) +
    xlab("Award amount ($)") +
    ylab("Total relative citation ratio")
```
</details>

```{r, echo = FALSE}
plotly::ggplotly(plot)
```

# Session information

```{r}
sessionInfo()
```
