---
title: "Case study: ITCR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: ITCR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<!-- badges: start -->
<!-- badges: end -->

Compiled: `r Sys.Date()`

# Summary

Note that this is way outside my area of expertise, so I have
undoubtedly made terrible blunders.

The 'grantpubcite' package can be used to query the [NIH Reporter][]
database for funded grants, and the publications associated with those
grants. The citation history of publications can be discovered using
[iCite][].

As a case study, suppose we are interested in grants funded under the
Information Technology in Cancer Research ([ITCR][]) program.

[ITCR]: https://itcr.cancer.gov/

The [Project information][] section shows that I found 146 projects (a
useful sanity check?). The first figure shows that projects funded for
more fiscal years received more funding (not too surprising!). At this
stage and subsequently I found it useful, using the datatables on the
web page, to search for ‘my’ project (U24CA180996, Cancer Genomics:
Integrative and Scalable Solutions in R/Bioconductor) and compare how
we were doing relative to other projects. Frankly, I did not know
where we stood; now I do.
 
I then focused on projects receiving funding in 2020.
 
The [Maturing projects?][] subsection uses project title and contact
PI to ask whether projects graduate between funding activities. One
(Campbell) has matured from 'Innovative' to 'Early-Stage', three
(Griffith, Karchin, and Liu) have matured from ‘Early-stage’ to
‘Advanced’, and two have matured from 'Advanced' to 'Sustained'. I was
a little disappointed that there were not more smaller projects
maturing.
 
In the [Publications][] section, we see a positive relationship
between funding (amount or duration) and publication. Again this is
not surprising, but speaks to the notion that scientific software
projects are playing by the same rules as ‘wet lab’ projects; it would
be interesting to identify a comparable wet-lab program for
comparison. There are some outliers, including one project with a
surprisingly large number of publications for the period of the grant,
as well as projects with only a handful of publications even after
multiple years of funding. The [ITCR collaboration][] section shows
that 15 projects collaborated closely enough with each other to be
acknowledged in the same publication.
 
The figure in the [Citations][] section shows what I imagine is a
pretty typical pattern, with most publications having moderate (<100)
citations, and a few having many citations. Scanning the datatable
immediately below the figure indicates that the highly cited
publications are biologically driven, where the paper authors have
presumably cited the relevant software; the three most-cited
publications are all from U24CA180922, which funds the Trinity
software. When I look at the five most highly cited of ‘my’ project
publications (U24CA180996), they are more modestly successful (181 to
337 citations); interestingly, 4 of the 5 publications emphasize
software per se, rather than strictly biological insight.
 
The [ITCR collaborations][] subsection takes a second look at
collaboration, in the sense of ITCR projects citing the works of one
another. 38 projects cite work of 30 other projects.

[NIH Reporter]: https://reporter.nih.gov/
[iCite]: https://icite.od.nih.gov/

# Getting started

See the [Introduction to 'grantpubcite'][grantpubcite] article for
installation, basic use, and a brief introduction to 'tidyverse'
operations.

[grantpubcite]: ./grantpubcite.html

Load the library and other packages to be used in this article.

<details>
```{r setup, message = FALSE}
library(grantpubcite)
library(ggplot2) # also plotly::ggplotly()
library(DT)
library(visNetwork)
```
</details>

This document is written in Rmarkdown; code chunks used to generate
each table or figure can be shown by toggling the 'Details' widget.

# Project information

The relevant funding opportunity announcements (FOA) are as follows.

<details>
```{r}
foas <- tribble(
    ~full_foa,       ~reissue_of,     ~foa_tag,      ~description,
    ## current PARs
    "PAR-15-334",    NA,              "Innovative",  "Development of Innovative Informatics Methods and Algorithms for Cancer Research and Management (R21)",
    "PAR-15-332",    "PAR-12-288",    "Early-Stage", "Early-Stage Development of Informatics Technologies for Cancer Research and Management (U01)",
    "PAR-15-331",    "PAR-13-294",    "Advanced",    "Advanced Development of Informatics Technologies for Cancer Research and Management (U24)",
    "PAR-15-333",    NA,              "Sustained",   "Sustained Support for Informatics Resources for Cancer Research and Management (U24)",
    ## RFAs
    "RFA-CA-23-014", "RFA-CA-22-021", "Innovative", "Development of Innovative Informatics Methods and Algorithms for Cancer Research and Management (R21 Clinical Trial Optional)",
    "RFA-CA-23-015", "RFA-CA-22-022", "Early-Stage", "Early-Stage Development of Informatics Technologies for Cancer Research and Management  (U01 Clinical Trial Optional)",
    "RFA-CA-23-016", "RFA-CA-22-023", "Advanced", "Advanced Development of Informatics Technologies for Cancer Research and Management (U24 Clinical Trial Optional)",
    "RFA-CA-23-017", "RFA-CA-22-024", "Sustained", "Sustained Support for Informatics Technologies for Cancer Research and Management (U24 Clinical Trial Optional)",
    ## 
    "RFA-CA-22-021", "RFA-CA-21-013", "Innovative", "Development of Innovative Informatics Methods and Algorithms for Cancer Research and Management (R21 Clinical Trial Optional)",
    "RFA-CA-22-022", "RFA-CA-21-014", "Early-Stage", "Early-Stage Development of Informatics Technologies for Cancer Research and Management  (U01 Clinical Trial Optional)",
    "RFA-CA-22-023", "RFA-CA-21-015", "Advanced", "Advanced Development of Informatics Technologies for Cancer Research and Management (U24 Clinical Trial Optional)",
    "RFA-CA-22-024", "RFA-CA-21-016", "Sustained", "Sustained Support for Informatics Technologies for Cancer Research and Management (U24 Clinical Trial Optional)",
    ## 
    "RFA-CA-21-013", "RFA-CA-20-007", "Innovative", "Development of Innovative Informatics Methods and Algorithms for Cancer Research and Management (R21 Clinical Trial Optional)",
    "RFA-CA-21-014", "RFA-CA-20-008", "Early-Stage", "Early-Stage Development of Informatics Technologies for Cancer Research and Management  (U01 Clinical Trial Optional)",
    "RFA-CA-21-015", "RFA-CA-20-009", "Advanced", "Advanced Development of Informatics Technologies for Cancer Research and Management (U24 Clinical Trial Optional)",
    "RFA-CA-21-016", "RFA-CA-20-010", "Sustained", "Sustained Support for Informatics Technologies for Cancer Research and Management (U24 Clinical Trial Optional)",
    ## 
    "RFA-CA-20-008", "RFA-CA-19-039", "Early-Stage", "Early-Stage Development of Informatics Technologies for Cancer Research and Management  (U01 Clinical Trial Optional)",
    "RFA-CA-20-007", "RFA-CA-19-038", "Innovative", "Development of Innovative Informatics Methods and Algorithms for Cancer Research and Management (R21 Clinical Trial Optional)",
    "RFA-CA-20-009", "RFA-CA-19-040", "Advanced", "Advanced Development of Informatics Technologies for Cancer Research and Management (U24 Clinical Trial Optional)",
    "RFA-CA-20-010", "RFA-CA-19-041", "Sustained", "Sustained Support for Informatics Technologies for Cancer Research and Management (U24 Clinical Trial Optional)",
    ## 
    "RFA-CA-19-039", NA,              "Early-Stage", "Early-Stage Development of Informatics Technologies for Cancer Research and Management  (U01 Clinical Trial Optional)",
    "RFA-CA-19-038", NA,              "Innovative", "Development of Innovative Informatics Methods and Algorithms for Cancer Research and Management (R21 Clinical Trial Optional)",
    "RFA-CA-19-040", NA,              "Advanced", "Advanced Development of Informatics Technologies for Cancer Research and Management (U24 Clinical Trial Optional)",
    "RFA-CA-19-041", NA,              "Sustained", "Sustained Support for Informatics Technologies for Cancer Research and Management (U24 Clinical Trial Optional)",
    "RFA-CA-19-042", NA,              "Education", "Informatics Technology for Cancer Research Education Resource (UE5 Clinical Trial Not Allowed)",
    ## earlier PARs
    "PAR-13-294",    "PAR-12-287",    "Advanced",    "Advanced Development of Informatics Technology (U24)",
    "PAR-12-287",    NA,              "Advanced",    "Advanced Development of Informatics Technology (U24)",
    "PAR-12-288",    NA,              "Early-Stage", "Early-Stage Development of Informatics Technology (U01)"
    ## FIXME: add 'Revsion of...' FOAs
)
foas <-
    foas |>
    select(-reissue_of) |> # drop 'reissue_of' for clarity
    mutate(
        foa_tag = factor(
            foa_tag,
            levels = c(
                "Innovative", "Early-Stage", "Advanced", "Sustained",
                "Education"
            )
        )
    )
```
</details>

```{r, echo = FALSE}
foas |>
    gpc_datatable(truncate_columns = FALSE)
```

The NIH Reporter contains quite a bit of information about each
grant. Following the [Get started][] article, we query the
`reporter_projects()` endpoint, restricting fields included in the
return value to a few of particular interest.  Retrieve these fields
for all projects in NIH reporter associated with the funding
announcements of interest. Perform some minor data cleaning by
removing leading and trailing whitespace from contact PI names. The
result is a tidyverse `tibble`, and we use 'tidy' semantics to explore
the data. Each project returned by NIH Reporter, is associated with
funding over multiple fiscal years and mechanisms (e.g., through
administrative supplements).  A few projects changed names within the
same award; these have been standardized to the most-recent name.

[Get started]: ./grantpubcite.html

<details>
```{r}
projects <- program_projects(foas, by = "project")
```
</details>

The `r nrow(projects)` funded projects are (`contact_pi` is the most
recent)

```{r, echo = FALSE}
projects |>
    arrange(desc(fiscal_year)) |>
    gpc_datatable()
```

The number of projects funded by each type of FOA, and the total
amount allcoated to date, is

```{r, echo = FALSE}
projects |>
    left_join(foas, by = "full_foa") |>
    group_by(foa_tag) |>
    summarize(
        n = n(),
        amount = sum(award_amount),
        mean = as.integer(mean(award_amount)),
        median = as.integer(median(award_amount))
    ) |>
    gpc_datatable()
```

```{r, echo = FALSE}
projects |>
    left_join(foas, by = "full_foa") |>
    ## filter(foa_tag != "Education") |>
    group_by(foa_tag, core_project_num) |>
    summarize(amount = sum(award_amount), .groups = "drop") |>
    ggplot(aes(foa_tag, amount)) +
    geom_boxplot(outlier.color = NA) +
    geom_jitter() +
    scale_y_continuous(labels = scales::comma) +
    labs(x = "Type of FOA", y = "Amount awarded ($)")
```

The number of projects funded per fiscal year is

```{r, echo = FALSE}
projects |>
    distinct(full_foa, core_project_num, fiscal_year) |>
    count(years_of_funding = fiscal_year) |>
    gpc_datatable()
```

<!-- Three project numbers have been funded under more than one FOA. -->

<!-- ```{r, echo = FALSE} -->
<!-- program_projects(foas, by = "foa") |> ungroup() |> -->
<!--     filter(duplicated(project)) |> -->
<!--     select(project, contact_pi, title) |> -->
<!--     gpc_datatable() -->
<!-- ``` -->

The following summarizes project funding across years

<details>
```{r, echo = FALSE}
plot <-
    projects |>
    ggplot(aes(
        fiscal_year, award_amount,
        text = paste0(core_project_num, ": ", gpc_shorten(project_title))
    )) +
    geom_point() +
    geom_smooth(
        method = "lm", formula = y ~ x,
        inherit.aes = FALSE, aes(fiscal_year, award_amount)
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_x_continuous(breaks = seq(1, 10), minor_breaks = NULL) +
    labs(x = "Fiscal years of award", y = "Total funding ($)")
```
</details>

```{r, echo = FALSE}
plotly::ggplotly(plot)
```

## Maturing projects?

A unique aspect of the ITCR FOA structure is that it envisions
projects starting at one FOA and 'maturing' to the next FOA, e.g., an
`Early-Stage` U01 matures to an `Advanced` U24. It is not possible to
assess this by tracking a project number through full FOA, because the
project number changes with FOA. Are there projects with the same
contact PI that have received awards from different FOA tags?

<details>
```{r}
tagged_pis <-
    projects |>
    distinct(full_foa, project_title, contact_pi_name) |>
    left_join(foas, by = "full_foa") |>
    select(foa_tag, contact_pi_name) |>
    distinct()

maturing_pis <-
    tagged_pis |>
    count(contact_pi_name, sort = TRUE) |>
    filter(n > 1L) |>
    select(-n) |>
    left_join(
        projects |>
        distinct(full_foa, core_project_num, contact_pi_name, project_title),
        by = "contact_pi_name",
        multiple = "all"
    ) |>
    left_join(foas, by = "full_foa") |>
    select(contact_pi_name, project_title, foa_tag) |>
    arrange(contact_pi_name, project_title, foa_tag)
```
</details>

```{r, echo = FALSE}
maturing_pis |>
    gpc_datatable()
```

Direct inspection suggests the following transitions:

```{r maturing projects, echo = FALSE}
tribble(
    ~contact_pi_name, ~from,         ~to,
    "Campbell",       "Innovative",  "Early-Stage",
    "Griffith",       "Early-Stage", "Advanced",
    "Karchin",        "Early-Stage", "Advanced",
    "Liu",            "Early-Stage", "Advanced",
    "Marcus",         "Advanced",    "Sustained",
    "Mesirov",        "Advanced",    "Sustained"
) |>
    gpc_datatable()
## FIXME: Kavraki, Narzisi?
```

# Publications

Grantees report publications associated with their grants, and this
information can be retrieved from NIH reporter. Queries are formulated
in a way similar to projects, as described on the NIH Reporter
[publication search API][], a rich set of query criteria can be used,
but fields included in the return are strictly limited.

[publication search API]: https://api.reporter.nih.gov/#!/Search/Publications

<details>
```{r}
publications <- program_publications(foas)
## citations: remove duplicates due to collaboration
citations <-
    publications |>
    select(-c("full_foa", "core_project_num")) |>
    distinct() |>
    select(
        "pmid", "citation_count", "relative_citation_ratio",
        "field_citation_rate", everything()
    ) |>
        arrange(desc(citation_count))
```
</details>

`r publications |> distinct(core_project_num) |> NROW()` projects
produced `r publications |> distinct(pmid) |> NROW()` publications.

Projects and their associated publications and citation statistics are
as follows; publications acknowledging more than one grant appear for
each acknowledgment.  with these projects are obtained with

```{r, echo = FALSE}
publications |>
    gpc_datatable()
```

## Project publication and funding

The most prolific projects are

<details>
```{r}
publications_by_project <-
    publications |>
    count(core_project_num, sort = TRUE, name = "n_publ") |>
    left_join(
        projects |>
        select(core_project_num, contact_pi_name, project_title, award_amount),
        by = "core_project_num"
    )
```
</details>

```{r, echo = FALSE}
publications_by_project |>
    gpc_datatable()
```

The large number of publications reported by projects 'U01CA239055'
and 'U01CA248226' are apparently because the software tools developed
by the group were immediately useful in completing projects at the
image analysis center. The relationship between publication and
funding amount are visualized as follows.

<details>
```{r}
plot <-
    publications_by_project |>
    ggplot(aes(
        award_amount, n_publ,
        text = paste0(core_project_num, ": ", gpc_shorten(project_title))
    )) +
    geom_point() +
    geom_smooth(
        method = "lm", formula = y ~ x,
        inherit.aes = FALSE, aes(award_amount, n_publ)
    ) +
    scale_x_continuous(labels = scales::comma) +
    labs(x = "Total funding ($)", y = "Number of publications")
```
</details>

```{r, echo = FALSE}
plotly::ggplotly(plot)
```

## ITCR collaboration

ITCR emphasizes collaboration between funded projects. Are there
examples of collaboration at the level of publication, i.e., `pmid`
associated with more than one project number?

<details>
```{r}
collaborative_publications <-
    publications |>
    count(pmid, sort = TRUE, name = "n_collab") |>
    filter(n_collab > 1) |>
    left_join(
        publications |>
        select(-c("full_foa", "core_project_num")) |>
        distinct(),
        by = "pmid"
    )
```
</details>

```{r, echo = FALSE}
collaborative_publications |>
    gpc_datatable()
```

Which projects are collaborating through shared publication?

<details>
```{r}
collaborative_projects <-
    copublication(foas) |>
    left_join(
        projects |> select(core_project_num, contact_pi_name, project_title),
        by = "core_project_num"
    ) |>
    arrange(desc(collab))
```
</details>

```{r, echo = FALSE}
collaborative_projects |>
    gpc_datatable()
```

Relationships are visualized as a network. Hover over nodes to see
project number and title. The width of edges is proportional to the
square root of the number of co-publications; dashed lines indicate a
single copublication. The strongest relationships are
'self-copublication' between grants from the same PI or institution.

<details>
```{r}
copub_data <- copublication_data(foas)
nodes <-
    copub_data |>
    tidyr::pivot_longer(dplyr::starts_with("core_project_num")) |>
    distinct(id = value) |>
    left_join(
        projects |> select(id = "core_project_num", project_title) |> distinct(),
        by = "id"
    ) |>
    mutate(
        size = 10,
        title = paste0(id, ": ", .data$project_title)
    )

edges <-
    copub_data |>
    mutate(
        from = core_project_num.x,
        to = core_project_num.y,
        width = 3 * sqrt(n),
        smooth = FALSE,
        dashes = n == 1L
    )

copub_network <-
    visNetwork(nodes, edges) |>
    ## set random seed for reproducibility
    visLayout(randomSeed = 123)
```
</details>

```{r, echo = FALSE}
copub_network
```

# Citations

Publications and their citation statistics are as follows:

```{r, echo = FALSE}
citations |>
    gpc_datatable()
```

The `r NROW(citations)` publications have 
`r summarize(citations, n = sum(citation_count)) |> pull(n)` total
citations; `r filter(citations, citation_count == 0) |> NROW()`
publications have not been cited; not surprisingly uncited
publications are recent.

Highly cited publications are in high-impact journals, and emphasize
science-related results rather than a software tool *per se* -- the
tool has been used in an important study, and the authors of the study
have acknowledged the tool.

Citations follow a very familiar pattern, with a few publications
cited frequently.

```{r, echo = FALSE}
citations |>
    ggplot(aes(x = 1 + citation_count)) +
    scale_x_log10() +
    geom_density() +
    xlab("Number of citations + 1") + ylab("Number of publications")
```

Citations per project are summarized below.

<details>
```{r, echo = FALSE}
project_summary <-
    projects |>
    select(
        core_project_num, contact_pi_name, project_title,
        fiscal_year, award_amount
    )

project_citations <-
    publications |>
    group_by(core_project_num) |>
    summarize(
        n_pub = n(),
        n_uncite = sum(citation_count == 0L, na.rm = TRUE),
        citation_count = sum(citation_count),
        relative_citation_ratio = sum(relative_citation_ratio, na.rm = TRUE)
    ) |>
    arrange(desc(citation_count)) |>
    left_join(project_summary, by = "core_project_num")
```
</details>

```{r, echo = FALSE}
project_citations |>
    gpc_datatable()
```

## ITCR collaborations

An opportunity for (indirect) collaboration occurs when one project
cites the work of another project. Thus we query [iCite][] for the
publications that cited ITCR publications, and exclude publications
that are not themselves ITCR publications.

<details>
```{r}
cocite_data <- cocitation_data(foas)

## ITCR publications ...
cocite_data |>
    distinct(pmid) |>
    NROW()
## ... cited by other ITCR publications
cocite_data |>
    distinct(cited_by) |>
    NROW()
## ITCR projects
cocite_data |>
    distinct(core_project_num) |>
    NROW()
## ... cited by other ITCR projects
cocite_data |>
    distinct(cited_by_core_project_num) |>
    NROW()

## pmid / cited_by pairs
total_citations <- NROW(cocite_data)

## self-citations
core_project_num_self_citations <-
    cocite_data |>
    count(self_citation = core_project_num == cited_by_core_project_num) |>
    filter(self_citation) |>
    pull(n)
```
</details>

There are
`r cocite_data |> distinct(pmid) |> NROW()` ITCR publications cited by
`r cocite_data |> distinct(cited_by) |> NROW()` ITCR publications.
The publications are from
`r cocite_data |> distinct(core_project_num) |> NROW()` ITCR projects cited by
`r cocite_data |> distinct(cited_by_core_project_num) |> NROW()` ITCR projects.

Of the total `r total_citations` pmid / citation pairs, 
`r core_project_num_self_citations` are projects citing their own
project.

<details>
```{r}
cocitation <-
    cocitation(foas) |>
    arrange(desc(n_self_citn))
```
</details>

```{r, echo = FALSE}
cocitation |>
    gpc_datatable()
```

# Data download

All project information can be downloaded with the following table

```{r, echo = FALSE}
projects |>
    gpc_datatable_download()
```

Query [NIH Reporter][] and [iCite][] for all publications from all
projects, and for extended citation metrics.

```{r, echo = FALSE}
publications |>
    gpc_datatable_download()
```

Publication ITCR citation data from [iCite][] are

```{r, echo = FALSE}
cocite_data |>
    gpc_datatable_download()
```

# Session information

```{r}
sessionInfo()
```
