---
title: "Case study: IMAT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: IMAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

Compile: `r Sys.Date()`

Note that this is way outside my area of expertise, so I have
undoubtedly made terrible blunders.

# Introduction

The 'grantpubcite' package can be used to query the [NIH Reporter][]
database for funded grants, and the publications associated with those
grants. The citation history of publications can be discovered using
[iCite][].

[NIH Reporter]: https://reporter.nih.gov/
[iCite]: https://icite.od.nih.gov/

In this case study we look at the Innovative Molecular Analysis
Technologies ([IMAT][]) program aimed at the development andb
integration of novel and emerging technologies in the support of
cancer research, diagnosis, and treatment. We track funding
opportunities from 2012 onward.

[IMAT]: https://imat.cancer.gov/funding-opportunities/funding-overview

# Getting started

See the [Introduction to 'grantpubcite'][grantpubcite] article for
installation, basic use, and a brief introduction to 'tidyverse'
operations.

[grantpubcite]: ./grantpubcite.html

Load the library and other packages to be used in this case study.

```{r setup, message = FALSE}
library(grantpubcite)
library(ggplot2)
library(DT)
library(visNetwork)
```

This document is written in Rmarkdown; code chunks used to generate
each table or figure can be shown by toggling the 'Details' widget.

# Projects

Funding Opportunity Announcements for IMAT since 2012 are summarized as follows:

<details>
```{r imat-foas}
foas <- case_study_foa_imat()

foa_full_tag_count <-
    foas |>
    count(full_tag, name = "n_foa")
```
</details>

```{r echo = FALSE}
foa_full_tag_count
```

Discover projects funded through these FOAs by querying NIH Reporter,
and use the information returned by a the query to summarize the
projects.

<details>
```{r}
project_summary <- program_projects(foas)

project_foa_full_tag_count <-
    project_summary |>
    left_join(foas, by = "opportunity_number") |>
    count(full_tag, name = "n_project")

foa_awards <-
    left_join(foa_full_tag_count, project_foa_full_tag_count, by = "full_tag") |>
    mutate(n_project = ifelse(is.na(n_project), 0L, n_project))
```
</details>

There are `r nrow(project_summary)` projects.

```{r, echo = FALSE}
project_summary |>
    gpc_datatable()
```

Some FOAs have not yet funded projects.

```{r, echo = FALSE}
foa_awards |> gpc_datatable()
```

Award amounts are summarized below.

<details>
```{r}
plot <-
    project_summary |>
    left_join(foas, by = "opportunity_number") |>
    ggplot(aes(full_tag, award_amount)) +
    theme(axis.text.x = element_text(angle = 15, vjust = 1, hjust=1)) +
    scale_y_continuous(labels = scales::comma) +
    geom_boxplot(outlier.color = NA) +
    geom_jitter() +
    xlab("Funding type") + ylab("Amount ($)")
```
</details>

```{r, echo = FALSE}
plot
```

# Publications and citations

Publications associated with projects are discovered using [NIH
Reporter][]. Citations are from [iCite][].

<details>
```{r}
program_publications <- program_publications(foas)

publications <-
    program_publications |>
    select(-c("opportunity_number", "core_project_num")) |>
    distinct() |>
    arrange(desc(citation_count))

publications_per_project <-
    program_publications |>
    count(core_project_num, name = "n_pub") |>
    left_join(project_summary, by = "core_project_num") |>
    select(-c("opportunity_number", "project_title", "fiscal_year")) |>
    arrange(desc(n_pub))

collaboration_summary <-
    program_publications |>
    count(pmid, name = "n_collaborators") |>
    count(n_collaborators, name = "n_publications")
```
</details>

There are `r publications |> NROW()` publications. One challenge with
is that a large project (e.g., U54MD000538) may have been funded
throughf IMAT for only a small amount of money or short period of time
as a subproject. Nonetheless, publications reflect the publication of
the large project, including activities not funded by IMAT.

```{r, echo = FALSE, warning = FALSE}
publications |>
    gpc_datatable()
```

The figure below shows publications per year...

<details>
```{r}
plot <-
    publications |>
    ggplot(aes(factor(year))) +
    geom_bar() +
    xlab("Year") + ylab("Number of publications")
```
</details>

```{r, echo = FALSE}
plot
```

...and the distribution of citations per publication

<details>
```{r}
plot <-
    publications |>
    filter(citation_count > 0) |>
    ggplot(aes(citation_count)) +
    scale_x_log10() +
    geom_density() +
    xlab("Number of citations") + ylab("Number of publications")
```
</details>

```{r, echo = FALSE}
plot
```


Project-level citations are summarized by the number of publications,
the total citation count, and the sum of the 'relative citation
index', a measure provided by iCite standardizing the impact of
publications by year and field of study.

<details>
```{r}
citations_by_project <-
    program_publications |>
    group_by(core_project_num) |>
    summarize(
        n_pub = n(),
        n_citn = sum(citation_count),
        total_rcr = sum(relative_citation_ratio, na.rm = TRUE)
    ) |>
    left_join(project_summary, by = "core_project_num") |>
    arrange(desc(total_rcr))
```
</details>

```{r, echo = FALSE}
citations_by_project |>
    gpc_datatable()
```

## Collaboration

The network below show collaboration between projects, as reflected in
publications acknowledging more than one project. Hover over nodes to
see project number and title. The width of edges is proportional to
the square root of the number of co-publications, dashed lines
indicate a single copublication; hover over edges to see the number of
collaborations.

<details>
```{r}
## pubs <- program_publications(foas)
## pubs |> count(pmid) # pmid 32554617 is highly collaborative
copub_data <- copublication_data(foas, exclude = "32554617")

nodes <-
    copub_data |>
    ## exclude
    tidyr::pivot_longer(dplyr::starts_with("core_project_num")) |>
    distinct(id = value) |>
    left_join(
        project_summary |>
        select(id = "core_project_num", project_title) |>
        distinct(),
        by = "id"
    ) |>
    mutate(
        size = 10,
        title = paste0(id, ": ", .data$project_title)
    ) |>
    arrange(id)

edges <-
    copub_data |>
    mutate(
        from = core_project_num.x,
        to = core_project_num.y,
        width = 3 * sqrt(n),
        smooth = FALSE,
        dashes = n == 1L,
        title = n
    )

network <-
    visNetwork(nodes, edges) |>
    visLayout(randomSeed = 123) |>
    visOptions(highlightNearest = TRUE)
```
</details>

```{r, echo = FALSE}
network
```

# Session information

```{r}
sessionInfo()
```
