---
title: "Case study: IOTN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: IOTN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Compiled: `r Sys.Date()`

Note that this is way outside my area of expertise, so I have
undoubtedly made terrible blunders.

# Introduction

The 'grantpubcite' package can be used to query the [NIH Reporter][]
database for funded grants, and the publications associated with those
grants. The citation history of publications can be discovered using
[iCite][].

[NIH Reporter]: https://reporter.nih.gov/
[iCite]: https://icite.od.nih.gov/

In this case study we look at the Immuno-Oncology Translational
Network ([IOTN][]), a consortium supported by the [NIH Cancer Moonshot][]
to accelerate translation of basic discoveries to clinical
applications to improve immunotherapy outcomes.

[IOTN]: https://www.iotnmoonshot.org/en/
[NIH Cancer Moonshot]: https://www.cancer.gov/research/key-initiatives/moonshot-cancer-initiative

# Getting started

See the [Introduction to 'grantpubcite'][grantpubcite] article for
installation, basic use, and a brief introduction to 'tidyverse'
operations.

[grantpubcite]: ./grantpubcite.html

Load the library and other packages to be used in this case study.

```{r setup, message = FALSE}
library(grantpubcite)
library(ggplot2)
library(DT)
```

This document is written in Rmarkdown; code chunks used to generate
each table or figure can be shown by toggling the 'Details' widget.

# Projects

Funding Opportunity Announcements for IOTN are as follows:

<details>
```{r}
foas <- tribble(
    ~full_foa,       ~tag,      ~description,
    "RFA-CA-19-044", "U01",     "Advancing Cancer Immunotherapy by Mitigating Immune-Related Adverse Events (irAE) (U01 Clinical Trial Not Allowed)",

    "RFA-CA-19-012", "UG3/UH3", "Immuno-Oncology Translation Network (IOTN): Cancer Immunoprevention Research Projects (UG3/UH3 Clinical Trials Not Allowed)",
    "RFA-CA-19-013", "U54",     "Immuno-Oncology Translation Network (IOTN): Immuno-engineering to Improve Immunotherapy (i3) Centers (U54 Clinical Trial Not Allowed)",
    "RFA-CA-19-014", "U01",     "Immuno-Oncology Translation Network (IOTN): Cancer Immunoprevention Research Projects (U01 Clinical Trial Not Allowed)",
    "RFA-CA-19-015", "U01",     "Immuno-Oncology Translation Network (IOTN): Cancer Immunotherapy Research Projects (U01 Clinical Trial Not Allowed)",

    "RFA-CA-17-045", "U01",     "Immuno-Oncology Translation Network (IOTN): Cancer Immunotherapy Research Projects (U01)",
    "RFA-CA-17-046", "U01",     "Immuno-Oncology Translation Network (IOTN): Cancer Immunoprevention Research Projects (U01)",
    "RFA-CA-17-047", "DMRC",    "Immuno-Oncology Translation Network (IOTN): Data Management and Resource-Sharing Center (DMRC) (U24)",
    "RFA-CA-17-048", "CIDR",    "Immuno-Oncology Translation Network (IOTN): Cellular Immunotherapy Data Resource (CIDR) (U24)"
) |>
    mutate(
        tag = factor(tag, levels = c("UG3/UH3", "U01", "DMRC", "CIDR", "U54"))
    )
```
</details>

```{r, echo = FALSE}
gpc_datatable(foas)
```

Discover projects funded through these FOAs by querying NIH Reporter,
and use the information returned by a the query to summarize the
projects.

<details>
```{r}
include_fields <- c(
    "full_foa", "core_project_num",
    "fiscal_year", "award_amount",
    "contact_pi_name", "project_title",
    "project_start_date", "project_end_date"
)
projects <- reporter_projects(
    foa = foas$full_foa,
    include_fields = include_fields
)

project_summary <-
    projects |>
    ## hack to place overall project ahead of 'cores' in program
    ## project grants
    arrange(desc(nchar(project_title))) |>
    group_by(full_foa, core_project_num) |>
    summarize(
        start_date = min(project_start_date, na.rm = TRUE),
        end_date = max(project_end_date, na.rm = TRUE),
        amount = sum(award_amount) |> as.integer(),
        contact_pi_name = head(contact_pi_name, 1L),
        project_title = head(project_title, 1L)
    ) |>
    left_join(foas |> select(full_foa, tag), by = "full_foa") |>
    select(full_foa, tag, everything())
```
</details>

```{r, echo = FALSE}
project_summary |>
    gpc_datatable()
```

There are `r nrow(project_summary)` projects. The projects include
large U54, resource sharing and management centers, as well as U01 and
UG3 awards.

```{r, echo = FALSE}
project_summary |>
    ggplot(aes(tag, amount)) +
    scale_y_continuous(labels = scales::comma) +
    geom_boxplot(outlier.color = NA) + geom_jitter() + coord_flip() +
    xlab("Funding type") + ylab("Amount ($)")
```

# Publications

Publications associated with projects are discovered using NIH Reporter.

<details>
```{r}
publications <- reporter_publications(
    core_project_nums = pull(project_summary, "core_project_num")
) |>
    select(-applid)
```
<details>

There are `r publications |> select(pmid) |> distinct() |> NROW()`
publications.

```{r, echo = FALSE}
publications |>
    count(coreproject, name = "n_pub") |>
    left_join(project_summary, by = c(coreproject = "core_project_num")) |>
    select(-project_title) |>
    gpc_datatable()
```

About 10% of publications involved collaboration between projects; one
publication represented an extensive collaboration.

```{r}
publications |>
    count(pmid, name = "n_collaborators") |>
    count(n_collaborators, name = "n_publications")
```

# Citations

Citations to each publication are obtained using `icite()`.

<details>
```{r}
include_fields <- c(
    "pmid", "year", "citation_count", "relative_citation_ratio",
    "year", "title", "authors", "journal", "doi"
)
citations <-
    icite(publications, include_fields = include_fields) |>
    distinct()

citations <- 
    left_join(publications, citations, by = "pmid") |>
    arrange(desc(citation_count))
```
</details>

```{r, echo = FALSE}
citations |>
    ## don't display 'author' field and truncate publication title;
    ## long fields are not handled well by datatable
    select( -authors ) |>
    mutate(title = paste0(substring(title, 1, 20), "...")) |>
    rename(
        n_citn = "citation_count",
        rcr = "relative_citation_ratio"
    ) |>
    gpc_datatable()
```

Project U01DK124165 had a very important publication with PMID 32839624. The
figure below shows publications per year...

```{r, echo = FALSE}
citations |>
    ggplot(aes(factor(year))) +
    geom_bar() +
    xlab("Year") + ylab("Number of publications")
```

...and the distribution of citations per publication

```{r}
citations |>
    filter(citation_count > 0) |>
    ggplot(aes(citation_count)) +
    scale_x_log10() +
    geom_density() +
    xlab("Number of citations") + ylab("Number of publications")
```

Project-level citations are summarized by the number of publications,
the total citation count, and the sum of the 'relative citation
index', a measure provided by iCite standardizing the impact of
publications by year and field of study.

<details>
```{r}
citations_by_project <-
    citations |>
    group_by(coreproject) |>
    summarize(
        n_pub = n(),
        n_citn = sum(citation_count),
        total_rcr = sum(relative_citation_ratio, na.rm = TRUE)
    ) |>
    left_join(project_summary, by = c(coreproject = "core_project_num")) |>
    arrange(desc(total_rcr))
```
</details>

```{r, echo = FALSE}
citations_by_project |>
    gpc_datatable()
```

The relationship between amount of funding and relative citation ratio
is show in the figure below; mousing over points shows the underlying
data and associated core project number. Large 'U54' projects do not
emphasize publication.

```{r, echo = FALSE}
plot <-
    citations_by_project |>
    mutate(`Start date` = format(start_date, "%Y")) |>
    ggplot(
        aes(
            amount, total_rcr,
            label = coreproject,
            color = `Start date`
        )
    ) +
    geom_point() +
    xlab("Award amount ($)") +
    ylab("Total relative citation ratio")

plotly::ggplotly(plot)
```

# Session information

```{r}
sessionInfo()
```
